\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}

This book is an experiment.

Category theory has a reputation: it is something you learn \emph{after} years
of mathematics, as a kind of universal solvent that dissolves the walls between
algebra, geometry, logic, and the rest. Every existing textbook on the subject
assumes you already live in those rooms.

This book tries something different. It treats category theory not as a
retrospective unifier of mathematics but as a \emph{starting point}. We build
from nothing. The only raw material is the idea that there are \term{things},
and between things there are \term{connections}. Everything else grows from
there. The goal is a full formal understanding of category theory — its
definitions, its theorems, its classical results — and eventually its
connections to other branches of mathematics. Those connections appear only
after we have built the language to appreciate them.

\bigskip

\textbf{Who this book is for.} If you can follow a recipe, read a transit map,
or understand why two different roads can take you to the same place, you have
all the prerequisites. We will not assume familiarity with algebra, calculus,
set theory, or any other branch of mathematics. When those subjects appear
later as examples, they will be introduced fresh.

\bigskip

\textbf{How each section works.} Each section follows a cycle. The cycle has
a direction: it opens outward from an intuitive starting point, and then folds
back inward to a precise formal statement. Different sections need different
numbers of steps, but the vocabulary of steps is always drawn from the same
small set:

\medskip
\begin{itemize}
  \item \textbf{Informally} ---
    The intuitive, everyday-language version of the idea. Start here every
    time.

  \item \textbf{Expanding} ---
    Worked examples, diagrams, and the kind of out-loud thinking you should
    practice when you encounter these ideas in the wild.

  \item \textbf{Formalizing} ---
    Precise definitions and notation are introduced. This is where the
    informal idea gets its mathematical clothing.

  \item \textbf{Reorganizing} ---
    We go back to the examples from the Expanding step and restate them
    \emph{using} the formal vocabulary we just learned. The facts are the
    same; the language is new.

  \item \textbf{Simplifying} ---
    We compress. Informal sentences are replaced, one by one, with their
    formal equivalents, until the statement fits in a line or two of notation.

  \item \textbf{Formally} ---
    The CT-idiomatic form. What a working category theorist would write, and
    what a reader who has completed the cycle can now read fluently.
\end{itemize}
\medskip

The cycle is not a secret. We use it openly, and we label each step so you
always know where you are. Not every section needs all six steps. What every
section does is complete a round trip: from plain language to formal notation
and back to a compressed formal statement that carries the full informal
meaning inside it.

\bigskip

\textbf{The internal-dialog boxes.} Throughout the book you will find boxes
labeled ``Internal Dialog.'' These model the running commentary that a skilled
reader maintains while working with diagrams and definitions. Learning to
produce that commentary yourself---in your own head, in pencil in the
margin---is one of the main skills this book tries to teach.

\bigskip

\textbf{Chapter bookends.} When a chapter covers several distinct ideas, it
opens with an \textbf{Overview} and closes with a \textbf{Formal Restatement}.
The Overview is where you are \emph{before} the chapter: a plain-language
description of what is coming. The Formal Restatement is where you are
\emph{after}: the same ideas, in CT-idiomatic notation. A reader who finishes
the chapter and reads the Formal Restatement should be able to hear the
Overview behind every line.

\bigskip

\textbf{Restarting is part of reading this book.} The cycle structure is
designed for repeated passes. On a first reading, you will use the Informal
and Expanding steps to build intuition. On a second pass, the Reorganizing
and Simplifying steps will feel different --- you will recognize the formal
language as yours. On a third pass, you may find yourself reading the Formal
Restatement first and working backwards. All of these are the right way to
read. The book does not penalize you for returning to earlier sections; it
rewards it.

\bigskip

\textbf{A note on what we avoid.} We will not say things like ``consider the
category of groups'' or ``the fundamental groupoid of a topological space''
until we have built those ideas from scratch. The temptation to borrow
ready-made examples from other fields is real, but it defeats the purpose. Our
examples come from everyday experience: maps and routes, instructions and steps,
classifications and comparisons. The mathematics, when it appears, will be
mathematics we have built together.

\bigskip

Let's begin.
