\chapter{Things and Arrows}

\begin{overviewbox}
Category theory is the study of \emph{things} and the \emph{connections between
things}. In this chapter we give those words precise names, examine several
everyday examples, and establish three rules that connections must follow:
they can be chained end-to-end; every thing connects to itself by a special
``do nothing'' connection; and when a diagram shows multiple paths between the
same two points, all paths must agree.

Each of those sentences is informal. By the end of the chapter each one
appears again in the compact notation a category theorist would write. The
Formal Restatement at the close of the chapter collects them together. A
reader returning to this chapter a second time may find it useful to read the
Formal Restatement first, then work through the chapter as a commentary on it.
\end{overviewbox}

% ═══════════════════════════════════════════════════════════════════════════
\section{What Category Theory Is About}
\label{sec:what-ct-is-about}
% ═══════════════════════════════════════════════════════════════════════════

\InformalLabel
\subsection{Informally: The Central Idea}

Category theory is the study of \term{things} and \term{arrows between things}.

That's it. That's the whole subject.

Of course, ``things'' and ``arrows'' are deliberately vague words. The power
of category theory comes precisely from that vagueness. A thing can be a
recipe step, a city, a sentence, a number, a proof, or a process. An arrow
can be a road, a translation, a logical implication, a comparison, or a
transformation of any kind.

What category theory cares about is not \emph{what things are made of inside},
but \emph{how they connect to other things}. Two things that connect to
everything else in exactly the same way are, for the purposes of category
theory, \emph{the same thing} --- even if they look completely different up
close.

This is a radical idea, and we will return to it many times.

% ─── Expanding ──────────────────────────────────────────────────────────────
\ExpandingLabel
\subsection{Expanding: Three Everyday Pictures}

Let's build the intuition with three examples that have nothing to do with
mathematics.

\begin{example}{The Transit Map}
A city has train stations. Between stations there are train lines. If you want
to get from Station~A to Station~C, and there is a line from A to B and a line
from B to C, you can travel A~$\to$~B~$\to$~C.

\medskip
\begin{center}
\begin{tikzcd}[column sep=2.5em]
  A \arrow[r, "\text{line 1}"] & B \arrow[r, "\text{line 2}"] & C
\end{tikzcd}
\end{center}
\medskip

The \emph{things} are the stations. The \emph{arrows} are the train lines.
Notice: we don't care what is inside a station --- its architecture, its size,
whether it has a coffee shop. We only care how it connects.
\end{example}

\begin{dialog}
When I see a diagram like the one above, I ask myself three questions:
\begin{enumerate}
  \item \emph{What are the dots?} Here: train stations.
  \item \emph{What are the arrows?} Here: train lines you can ride.
  \item \emph{Can I follow a path?} Here: yes, A to B to C.
\end{enumerate}
These three questions apply to \emph{every} diagram in this book. Make them
into a habit.
\end{dialog}

\begin{example}{The Recipe Steps}
A recipe is a sequence of steps. ``Chop onions'' must happen before ``sauté
onions.'' ``Sauté onions'' must happen before ``add to pot.''

\medskip
\begin{center}
\begin{tikzcd}[column sep=2.5em]
  \text{chop} \arrow[r, "\text{before}"] &
  \text{sauté} \arrow[r, "\text{before}"] &
  \text{add to pot}
\end{tikzcd}
\end{center}
\medskip

The \emph{things} are the steps. The \emph{arrows} mean ``must happen before.''
If chopping must happen before sautéing, and sautéing before adding, then
chopping must happen before adding --- even though there is no single arrow
drawn directly from ``chop'' to ``add to pot.''

This indirect conclusion is not magic. It follows from the arrows we have.
In category theory we will give this a formal name: \term{composition}.
\end{example}

\begin{example}{The Approval Chain}
A company requires that budget requests be approved by a manager before going
to the director, and by the director before going to the board.

\medskip
\begin{center}
\begin{tikzcd}[column sep=2.5em]
  \text{request} \arrow[r, "\text{manager}"] &
  \text{approved} \arrow[r, "\text{director}"] &
  \text{final approval}
\end{tikzcd}
\end{center}
\medskip

Again: \emph{things} (states of the request), \emph{arrows} (approvals that
move the request forward). The pattern is identical to the transit map and the
recipe. This is the first sign that something deeper is going on: very
different situations share the same shape.

Category theory is, among other things, the science of recognizing when two
situations have the same shape.
\end{example}

% ─── Formalizing ────────────────────────────────────────────────────────────
\FormalizingLabel
\subsection{Formalizing: Inside Versus Outside}

There is a habit of mind that category theory requires, and it is worth
practicing from the very beginning.

\textbf{Stop asking what things are made of.}

When you look at a train station on a map, you do not ask how many platforms
it has or which architect designed it. You ask: which lines does it connect to?
Category theory extends this discipline to \emph{everything}. When we work
with a thing in category theory, we are forbidden from looking ``inside'' it.
We can only look at what arrows arrive at it and what arrows leave it.

This sounds limiting. In practice it is liberating.

Suppose you have two completely different-looking objects --- a subway station
and an airport terminal, say --- but every route that goes through one also
goes through the other, and vice versa. For the purposes of your journey, they
are \emph{interchangeable}. Category theory gives you a way to say that
precisely and work with it rigorously.

\begin{readernote}
Try this: pick any two things in your life that are functionally
interchangeable but look different inside. A paperback and an e-book of the
same novel. An email and a printed letter carrying the same message. A
physical key and a key card opening the same door.

In each case: the \emph{inside} differs, but the \emph{connections} are the
same. In category theory we say these are \term{isomorphic} --- a word we will
define carefully later.
\end{readernote}

The discipline of ``look only at connections, not at insides'' is called
working \term{externally} or working \term{up to isomorphism}. Start training
the habit now: when you look at a thing, ask not \emph{what is it?} but
\emph{what does it connect to?}

\medskip

\textbf{A historical note.} Category theory was introduced in 1945 by Samuel
Eilenberg and Saunders Mac~Lane. Their language turned out to carry a
philosophical viewpoint: the question \emph{``what is X?''} is replaced by
\emph{``what role does X play?''} This is sometimes called \term{structuralism}
in the philosophy of mathematics. This book does not require you to adopt a
philosophical position, but it is worth knowing the viewpoint is there --- it
shapes every definition we write.


% ═══════════════════════════════════════════════════════════════════════════
\section{Objects and Morphisms}
\label{sec:objects-morphisms}
% ═══════════════════════════════════════════════════════════════════════════

\InformalLabel
\subsection{Informally: Naming the Pieces}

The \emph{things} in category theory have an official name: \term{objects}.

The \emph{arrows} have an official name: \term{morphisms}.

Every morphism goes \emph{from} one object \emph{to} another. We write
\[
  f : A \to B
\]
to mean ``there is a morphism named~$f$ going from object~$A$ to object~$B$.''
We call $A$ the \term{domain} of~$f$ and $B$ the \term{codomain} of~$f$.

Four words: objects, morphisms, domain, codomain. Everything else in category
theory is built from these four words.

% ─── Expanding ──────────────────────────────────────────────────────────────
\ExpandingLabel
\subsection{Expanding: Reading Morphism Notation}

Let's slow down and practice reading $f : A \to B$, because it will appear on
almost every page of this book.

\begin{dialog}
I see the expression $f : A \to B$. Here is what I say to myself:
\begin{itemize}
  \item ``$f$ is the name of a morphism.''
  \item ``$A$ is where $f$ starts --- its domain.''
  \item ``$B$ is where $f$ ends --- its codomain.''
  \item ``The arrow $\to$ tells me this is a one-directional connection.''
  \item ``I do \emph{not} assume I can go from $B$ back to $A$ just because
        I can go from $A$ to $B$. Arrows have directions.''
\end{itemize}
If someone asks me to draw this, I draw a dot labeled $A$, a dot labeled $B$,
and an arrow from $A$ to $B$ labeled $f$:
\medskip
\begin{center}
\begin{tikzcd}
  A \arrow[r, "f"] & B
\end{tikzcd}
\end{center}
\medskip
This picture and the expression $f : A \to B$ carry exactly the same
information.
\end{dialog}

\begin{example}{Morphisms as Temperature Comparisons}
Suppose our objects are temperature readings: $0°\text{C}$, $20°\text{C}$,
$37°\text{C}$, $100°\text{C}$ (freezing, room temperature, body temperature,
boiling). A morphism $f : 0°\text{C} \to 100°\text{C}$ could mean ``is colder
than.'' A morphism $g : 20°\text{C} \to 37°\text{C}$ could also mean ``is
colder than.''

Both $f$ and $g$ are instances of the same \emph{kind} of arrow (``is colder
than''), but they are different morphisms because they connect different
objects. Two morphisms are the same only if they have the same name \emph{and}
the same domain \emph{and} the same codomain.
\end{example}

\begin{example}{Morphisms as Translations}
Suppose our objects are the phrases ``hello,'' ``hola,'' ``bonjour,'' ``ciao.''
A morphism $t : \text{hello} \to \text{hola}$ could mean ``is a Spanish
translation of.'' A morphism $u : \text{hello} \to \text{bonjour}$ could mean
``is a French translation of.''

Both start at the same object (``hello'') but end at different objects. They
are different morphisms. The object ``hello'' has \emph{two outgoing morphisms}
in this setup.
\end{example}

\begin{readernote}
When you encounter any new example, before doing anything else, identify:
\begin{enumerate}
  \item What are the \emph{objects}? (the dots in the diagram)
  \item What are the \emph{morphisms}? (the arrows)
  \item For each morphism: what is its \emph{domain}? what is its
        \emph{codomain}?
\end{enumerate}
Write these down, in words, every time. This routine prevents most of the
confusion that comes from reading diagrams too quickly.
\end{readernote}

% ─── Formalizing ────────────────────────────────────────────────────────────
\FormalizingLabel
\subsection{Formalizing: What Morphisms Are Not}

Several natural assumptions about morphisms turn out to be wrong. Let's clear
them now.

\textbf{Morphisms are not always functions.}
In many examples, morphisms will be functions from one set to another. But
``morphism'' does not mean ``function.'' A morphism is anything that fits the
pattern: domain, codomain, goes from one to the other consistently. We will
see morphisms that are proofs, comparisons, paths, and transformations.

\textbf{Multiple morphisms can share the same domain and codomain.}
If $f : A \to B$ and $g : A \to B$, that does \emph{not} mean $f = g$.
There can be many different morphisms between the same two objects.

\textbf{A morphism $f : A \to B$ does not imply a morphism $B \to A$.}
Going from $A$ to $B$ is not the same as going from $B$ to $A$. In the recipe
example, ``chop before sauté'' does not imply ``sauté before chop'' --- that
would be a contradiction.

\textbf{There may be no morphism between two objects.}
Not every pair of objects is connected. Not every city has a direct train to
every other city.

% ─── Reorganizing ───────────────────────────────────────────────────────────
\ReorganizingLabel
\subsection{Reorganizing: The Earlier Examples in Formal Terms}

Now that we have the vocabulary, let's restate the examples from
Section~\ref{sec:what-ct-is-about} using it.

\textbf{Transit map.} The stations are \emph{objects}. The line from Station~A
to Station~B is a \emph{morphism} with domain~$A$ and codomain~$B$. ``There
is a train line from A to B'' becomes: there exists a morphism $f : A \to B$.
``There is no direct route from B to A'' becomes: there is no morphism with
domain~$B$ and codomain~$A$ in this example.

\textbf{Temperature.} The readings $0°\text{C}$, $20°\text{C}$,
$37°\text{C}$, $100°\text{C}$ are \emph{objects}. The relation ``is colder
than'' provides a \emph{morphism} $f : 0°\text{C} \to 100°\text{C}$. The
domain is the colder reading; the codomain is the warmer one.

\textbf{Translations.} ``Hello,'' ``hola,'' ``bonjour'' are \emph{objects}.
``Is a Spanish translation of'' is a \emph{morphism} $t : \text{hello} \to
\text{hola}$. A second Spanish translation (formal versus informal) would be
a second, distinct morphism $t' : \text{hello} \to \text{hola}$ --- same
domain, same codomain, but $t \neq t'$.

In every case, the informal description maps directly onto the formal
vocabulary. No facts changed; only the language did.

% ─── Simplifying ────────────────────────────────────────────────────────────
\SimplifyingLabel
\subsection{Simplifying: Four Phrases, Four Terms}

The informal sentences we used have formal counterparts:

\medskip
\begin{center}
\begin{tabular}{ll}
  \textit{Informal phrase} & \textit{Formal term} \\[4pt]
  \hline\\[-6pt]
  ``there is a connection from $X$ to $Y$'' & $f : X \to Y$ \\
  ``where the connection starts'' & domain \\
  ``where the connection ends'' & codomain \\
  ``all connections from $X$ to $Y$'' & $\mathrm{Hom}(X, Y)$ \\
\end{tabular}
\end{center}
\medskip

The last entry, $\mathrm{Hom}(X, Y)$, is the \term{hom-set}: the collection
of all morphisms with domain~$X$ and codomain~$Y$. We have replaced four
English phrases with four mathematical terms. The terms carry the same meaning
in any context.

% ─── Formally ───────────────────────────────────────────────────────────────
\FormallyLabel
\subsection{Formally: Morphisms and Hom-Sets}

A \term{morphism} $f : A \to B$ is determined by three data: its domain~$A$,
its codomain~$B$, and its identity as a morphism. Two morphisms are equal if
and only if they have equal domains, equal codomains, and are the same arrow.

The \term{hom-set} $\mathrm{Hom}(A, B)$ is the collection of all morphisms
with domain~$A$ and codomain~$B$. When we need to specify the category~$\mathcal{C}$,
we write $\mathrm{Hom}_{\mathcal{C}}(A, B)$.

\formalnote{Alternative notations}
In some literature: $f \in \mathrm{Hom}(A, B)$; or inline as
$A \xrightarrow{f} B$; or in diagrams without labels when the morphism is
clear from context. This book defaults to $f : A \to B$.

\formalnote{Hom-sets may be empty}
$\mathrm{Hom}(A, B) = \varnothing$ is permitted --- there need not be a
morphism between every pair of objects. The definition requires only that
\emph{when} morphisms exist, they obey the rules of composition and identity
(introduced in the next two sections).


% ═══════════════════════════════════════════════════════════════════════════
\section{Composition}
\label{sec:composition}
% ═══════════════════════════════════════════════════════════════════════════

\InformalLabel
\subsection{Informally: Chaining Arrows}

If there is a morphism from $A$ to $B$, and a morphism from $B$ to $C$, then
there must also be a morphism from $A$ to $C$.

This is the rule of \term{composition}. Whenever two arrows line up end-to-end,
you get a third arrow for free --- the one that goes from start to finish in one
step.

\medskip
\begin{center}
\begin{tikzcd}[column sep=3em, row sep=2em]
  A \arrow[r, "f"] \arrow[rr, bend right=25, "g \circ f"'] & B \arrow[r, "g"] & C
\end{tikzcd}
\end{center}
\medskip

We write the composed morphism as $g \circ f$, read ``$g$ after $f$.'' It goes
from $A$ to $C$. The order reads right to left: first $f$, then $g$.

% ─── Expanding ──────────────────────────────────────────────────────────────
\ExpandingLabel
\subsection{Expanding: Practicing Composition}

\begin{dialog}
I see: $f : A \to B$ and $g : B \to C$. I ask: ``Do these line up?'' The
codomain of~$f$ is~$B$. The domain of~$g$ is~$B$. They match. So composition
is possible.

I write: $g \circ f : A \to C$.

I read this as: ``first do $f$, then do $g$.'' I always verify:
domain of composite $=$ domain of $f = A$; codomain of composite $=$ codomain
of $g = C$.
\end{dialog}

\begin{example}{Composition in the Recipe}
Recall:
\[
  f : \text{chop} \to \text{sauté}, \qquad g : \text{sauté} \to \text{add to pot}
\]
Composition gives:
\[
  g \circ f : \text{chop} \to \text{add to pot}
\]
``Chopping must happen before adding to pot'' --- a fact we knew intuitively.
Composition is the formal way to record it.
\end{example}

\begin{example}{A Chain of Three}
Suppose $f : A \to B$, $g : B \to C$, $h : C \to D$. We can compose in steps:

\medskip
\begin{center}
\begin{tikzcd}[column sep=2.5em]
  A \arrow[r, "f"] \arrow[rrr, bend right=35, "h \circ g \circ f"']
  & B \arrow[r, "g"] \arrow[rr, bend left=30, "h \circ g"]
  & C \arrow[r, "h"]
  & D
\end{tikzcd}
\end{center}
\medskip

Both $(h \circ g) \circ f$ and $h \circ (g \circ f)$ give the same morphism
$A \to D$. This is not a coincidence --- it is a law called \term{associativity}.
\end{example}

% ─── Formalizing ────────────────────────────────────────────────────────────
\FormalizingLabel
\subsection{Formalizing: Associativity}

Associativity says that when composing a chain of three morphisms, the
grouping of intermediate steps doesn't matter:
\[
  h \circ (g \circ f) = (h \circ g) \circ f
\]

Why does this matter? Because it means we can write $h \circ g \circ f$
without parentheses --- the chain has a unique composite regardless of how we
group the steps.

\begin{example}{Associativity in Transit}
\begin{itemize}
  \item $f$: travel Airport $\to$ Downtown
  \item $g$: travel Downtown $\to$ Stadium
  \item $h$: travel Stadium $\to$ Hotel
\end{itemize}
Whether you plan the trip as (Airport$\to$Stadium) then Hotel, or Airport then
(Downtown$\to$Hotel), the journey is the same. Grouping the \emph{planning}
differently does not change the \emph{journey}. Associativity just says this
obvious fact always holds for morphisms.
\end{example}

Associativity has a useful consequence: any finite chain of composable arrows
has a unique composite. We don't need to specify parenthesization. The
composite of $f_1, f_2, \ldots, f_n$ (when they chain in order) is
well-defined.

Composition is also a \emph{partial} operation: $g \circ f$ is defined
\emph{only} when $\mathrm{cod}(f) = \mathrm{dom}(g)$. The domain and codomain
are the type-checking mechanism that tells you which compositions are legal.

% ─── Reorganizing ───────────────────────────────────────────────────────────
\ReorganizingLabel
\subsection{Reorganizing: Recipe Composition in Formal Terms}

In the recipe example, $f : \text{chop} \to \text{sauté}$ and
$g : \text{sauté} \to \text{add}$. These are composable because
$\mathrm{cod}(f) = \mathrm{dom}(g) = \text{sauté}$.

Their composite $g \circ f : \text{chop} \to \text{add}$ is the formal
encoding of ``chopping must happen before adding.'' The intermediate step
(sautéing) is \emph{inside} the composite; it doesn't appear in the domain
or codomain of $g \circ f$.

For the transit chain --- Airport, Downtown, Stadium, Hotel --- we have
morphisms $f$, $g$, $h$ and composites $g \circ f$, $h \circ g$, and
$h \circ g \circ f$. Associativity says the last one is unambiguous:
$(h \circ g) \circ f = h \circ (g \circ f) = h \circ g \circ f$.

% ─── Simplifying ────────────────────────────────────────────────────────────
\SimplifyingLabel
\subsection{Simplifying: The Composition Rule}

The entire informal discussion of chaining compresses to two lines:
\[
  f : A \to B,\quad g : B \to C \implies g \circ f : A \to C
\]
\[
  (h \circ g) \circ f = h \circ (g \circ f)
\]

The first line says when composition is defined and what it produces. The
second line says that the order of grouping doesn't matter. Everything else
we said was commentary on these two lines.

% ─── Formally ───────────────────────────────────────────────────────────────
\FormallyLabel
\subsection{Formally: Composition as a Partial Operation}

Composition is a family of functions
\[
  \circ : \mathrm{Hom}(B, C) \times \mathrm{Hom}(A, B) \;\to\; \mathrm{Hom}(A, C)
\]
one for each triple of objects $A$, $B$, $C$. It satisfies \term{associativity}:
for all $f \in \mathrm{Hom}(A,B)$, $g \in \mathrm{Hom}(B,C)$,
$h \in \mathrm{Hom}(C,D)$,
\[
  h \circ (g \circ f) = (h \circ g) \circ f.
\]

This is equivalently expressed as a commutative diagram: the two composite
paths from $A$ to $D$ in the chain $A \to B \to C \to D$ are equal.

\formalnote{Why composition precedes identity in this presentation}
Composition is the operative structure of a category. Identity morphisms
(next section) are a \emph{constraint} on composition: they are the neutral
elements that leave composites unchanged. Logically you must understand the
operation before you can understand its neutral element.


% ═══════════════════════════════════════════════════════════════════════════
\section{Identity Morphisms}
\label{sec:identity}
% ═══════════════════════════════════════════════════════════════════════════

\InformalLabel
\subsection{Informally: Every Object Has a ``Stay Put'' Arrow}

Every object has a special morphism that goes from \emph{itself to itself} and
does nothing. We call it the \term{identity morphism} on that object and write
$\mathrm{id}_A : A \to A$.

``Does nothing'' has a precise meaning: composing any morphism with the
identity gives back the original morphism unchanged. For any $f : A \to B$:
\[
  f \circ \mathrm{id}_A = f \qquad \text{and} \qquad \mathrm{id}_B \circ f = f
\]

\begin{center}
\begin{tikzcd}[column sep=3em]
  A \arrow[r, "f"] \arrow[loop left, "\mathrm{id}_A"]
  & B \arrow[loop right, "\mathrm{id}_B"]
\end{tikzcd}
\end{center}

% ─── Expanding ──────────────────────────────────────────────────────────────
\ExpandingLabel
\subsection{Expanding: The ``Do Nothing'' Connection}

\begin{example}{The Optional Rest Step}
Suppose a recipe optionally rests dough for an hour. On a day when you're in
a hurry, you skip the step: the dough is the same dough it was before.
This ``do nothing'' is an identity morphism: it takes the dough as input and
returns it unchanged.
\end{example}

\begin{example}{The Zero-Distance Route}
In the transit example, ``stay at Station A'' is a valid route: you start at A
and end at A. It composes correctly with other routes:
\[
  (\text{stay at A}) \text{ then } (\text{A to B}) = (\text{A to B})
\]
\[
  (\text{A to B}) \text{ then } (\text{stay at B}) = (\text{A to B})
\]
The ``stay put'' route is the identity morphism on Station A.
\end{example}

\begin{dialog}
When I see a loop arrow on an object --- an arrow that starts and ends at
the same dot --- my first question is: ``Is this the identity, or a
non-trivial self-loop?''

I can tell them apart by testing: does composing with this arrow change any
other morphism? If no, it's the identity. If yes, it's something else.
\end{dialog}

\textbf{Uniqueness of identity.} Each object has \emph{exactly one} identity
morphism. Suppose $\mathrm{id}_A$ and $\mathrm{id}_A'$ both satisfy the
neutrality law. Then:
\[
  \mathrm{id}_A
  = \mathrm{id}_A \circ \mathrm{id}_A'
  = \mathrm{id}_A'
\]
The first equality uses $\mathrm{id}_A'$ as a right identity; the second uses
$\mathrm{id}_A$ as a left identity. They must be the same morphism.

\begin{readernote}
This is the first categorical proof in the book. Notice its structure: we
used the identity \emph{laws} (equations that must hold) rather than looking
inside any object. That is the categorical style. If a proof ever asks you to
look inside an object, something has gone wrong.
\end{readernote}

% ─── Formalizing ────────────────────────────────────────────────────────────
\FormalizingLabel
\subsection{Formalizing: Neutrality}

The identity morphism is characterized entirely by the equations it satisfies.
For any $f : A \to B$:
\[
  f \circ \mathrm{id}_A = f \qquad (\text{right unit law})
\]
\[
  \mathrm{id}_B \circ f = f \qquad (\text{left unit law})
\]

These are called the \term{unit laws}. Together they say: composing with the
identity on the appropriate side leaves any morphism unchanged.

% ─── Reorganizing ───────────────────────────────────────────────────────────
\ReorganizingLabel
\subsection{Reorganizing: Restating the Examples Formally}

\textbf{Transit.} The ``stay at Station A'' route is $\mathrm{id}_A \in
\mathrm{Hom}(A, A)$. For any route $f : A \to B$: $f \circ \mathrm{id}_A = f$
(riding the zero-distance route at A, then taking route $f$, is just route
$f$). And $\mathrm{id}_B \circ f = f$ (taking route $f$ and then staying at B
is just route $f$).

\textbf{Recipe.} The ``skip the rest step'' action is $\mathrm{id}_\text{dough}
\in \mathrm{Hom}(\text{dough}, \text{dough})$. Doing nothing before or after
any step leaves that step unchanged.

In both cases, the informal description maps exactly onto the unit laws.

% ─── Simplifying ────────────────────────────────────────────────────────────
\SimplifyingLabel
\subsection{Simplifying: Two Laws, One Line}

The two unit laws collapse to a single display:
\[
  f \circ \mathrm{id}_A = f = \mathrm{id}_B \circ f \qquad \text{for all } f : A \to B
\]

This single line captures everything the identity morphism does. The uniqueness
proof shows that $\mathrm{id}_A$ is the \emph{only} morphism satisfying this
condition.

% ─── Formally ───────────────────────────────────────────────────────────────
\FormallyLabel
\subsection{Formally: Identity as Part of the Data}

The identity is not derived --- it is specified. Part of the data of a category
is an assignment: to each object $A$, a distinguished morphism
$\mathrm{id}_A \in \mathrm{Hom}(A, A)$, satisfying:
\[
  f \circ \mathrm{id}_A = f = \mathrm{id}_B \circ f \qquad
  \text{for all } A, B, f : A \to B.
\]

The full category definition --- collecting objects, morphisms, composition,
and identity together --- is the subject of Chapter~2.

\formalnote{Displayed as triangles}
The unit laws are often displayed as commutative triangles:

\medskip
\begin{center}
\begin{tikzcd}[column sep=3em, row sep=2em]
  A \arrow[r, "f"] \arrow[dr, "f"']
    & B \arrow[d, "\mathrm{id}_B"] \\
  & B
\end{tikzcd}
\qquad
\begin{tikzcd}[column sep=3em, row sep=2em]
  A \arrow[r, "\mathrm{id}_A"] \arrow[dr, "f"']
    & A \arrow[d, "f"] \\
  & B
\end{tikzcd}
\end{center}
\medskip

Each diagram says: the two paths from the top-left object to the
bottom-right object are equal. That is precisely $\mathrm{id}_B \circ f = f$
and $f \circ \mathrm{id}_A = f$.


% ═══════════════════════════════════════════════════════════════════════════
\section{Commutative Diagrams}
\label{sec:commutative-diagrams}
% ═══════════════════════════════════════════════════════════════════════════

\InformalLabel
\subsection{Informally: When All Paths Agree}

A \term{commutative diagram} is a diagram of objects and morphisms in which
every path between the same two endpoints gives the same composite morphism.

The simplest example is a triangle:

\medskip
\begin{center}
\begin{tikzcd}[column sep=3em, row sep=2.5em]
  A \arrow[r, "f"] \arrow[dr, "h"'] & B \arrow[d, "g"] \\
  & C
\end{tikzcd}
\end{center}
\medskip

This diagram \emph{commutes} when $g \circ f = h$. Two paths from $A$ to $C$;
one composite result.

% ─── Expanding ──────────────────────────────────────────────────────────────
\ExpandingLabel
\subsection{Expanding: Reading Commutative Diagrams}

\begin{dialog}
I see a diagram with multiple paths between two objects. My procedure:
\begin{enumerate}
  \item \textbf{Identify} all paths from each object to each other object.
        I trace every possible route, like reading a map.
  \item \textbf{Write} the composite morphism for each path using $\circ$
        notation.
  \item \textbf{Ask}: are these composites equal? If yes, the diagram (or
        those paths) commutes.
  \item \textbf{Note}: a triangle asserts one equation, a square asserts one,
        a more complex diagram may assert many simultaneously.
\end{enumerate}
\end{dialog}

\begin{example}{A Commutative Square}
The most common diagram in category theory:

\medskip
\begin{center}
\begin{tikzcd}[column sep=3.5em, row sep=2.5em]
  A \arrow[r, "f"] \arrow[d, "p"'] & B \arrow[d, "q"] \\
  C \arrow[r, "g"'] & D
\end{tikzcd}
\end{center}
\medskip

This commutes when $q \circ f = g \circ p$: going right-then-down equals
going down-then-right.

\medskip

In everyday terms: suppose $A$ is a sentence in English, $B$ a formal logical
statement, $C$ the same sentence in French, $D$ the same formal statement.
Then $f$ = formalize English, $p$ = translate English to French, $q$ =
formalize French, $g$ = match formalizations. The square commutes if
``translate then formalize'' gives the same result as ``formalize then
translate.'' This is a genuine question, not a foregone conclusion.
\end{example}

\begin{example}{The Recipe as a Commutative Triangle}
We have $f : \text{chop} \to \text{sauté}$, $g : \text{sauté} \to \text{add}$,
and their composite $g \circ f : \text{chop} \to \text{add}$. If we also draw
$h : \text{chop} \to \text{add}$ meaning ``chop must happen before add,'' then
the triangle

\medskip
\begin{center}
\begin{tikzcd}[column sep=3em, row sep=2.5em]
  \text{chop} \arrow[r, "f"] \arrow[dr, "h"'] &
  \text{sauté} \arrow[d, "g"] \\
  & \text{add}
\end{tikzcd}
\end{center}
\medskip

commutes when $h = g \circ f$: the direct ``chop before add'' relation is
the same as the two-step route via sautéing. Commutativity here says these two
descriptions of the same ordering constraint are equal.
\end{example}

\begin{readernote}
Whenever you encounter a diagram, write next to it: ``This commutes because
\_\_\_\_\_\_.'' If you can't fill in the blank, you don't yet understand what
the diagram is asserting. Diagrams are \emph{claims} --- they say that two
composed morphisms are equal. Read them as equations, not just pictures.
\end{readernote}

% ─── Formalizing ────────────────────────────────────────────────────────────
\FormalizingLabel
\subsection{Formalizing: Diagrams as Equations}

Every commutative diagram is shorthand for a collection of equations between
composed morphisms. The diagram is often easier to parse than the equations;
the two are entirely equivalent.

A triangle asserts one equation. A square asserts one equation. A more complex
diagram with five or six objects may assert many equations simultaneously, one
for each pair of objects with multiple paths between them.

There is also a converse skill: given a collection of equations relating
composed morphisms, you can often \emph{draw} them as a commutative diagram.
Practicing this translation in both directions builds fluency.

Not every diagram in category theory is required to commute. Sometimes a
diagram merely depicts objects and morphisms without asserting equations.
Convention: when a diagram is stated to commute (explicitly or by context),
it does. When in doubt, check whether the author says ``the following diagram
commutes'' or just ``consider the following diagram.''

% ─── Reorganizing ───────────────────────────────────────────────────────────
\ReorganizingLabel
\subsection{Reorganizing: Diagrams and Their Equations}

\textbf{Triangle.} The triangle with $f : A \to B$, $g : B \to C$,
$h : A \to C$ commutes iff $h = g \circ f$. Equivalently: $h \in
\mathrm{Hom}(A, C)$ and $g \circ f \in \mathrm{Hom}(A, C)$ and they are
equal.

\textbf{Square.} The square with $f : A \to B$, $q : B \to D$, $p : A \to C$,
$g : C \to D$ commutes iff $q \circ f = g \circ p$ in $\mathrm{Hom}(A, D)$.

\textbf{Pasting.} If we have two commutative squares sharing an edge:

\medskip
\begin{center}
\begin{tikzcd}[column sep=3em, row sep=2.5em]
  A \arrow[r, "f"] \arrow[d, "p"']
  & B \arrow[r, "g"] \arrow[d, "q"]
  & C \arrow[d, "r"] \\
  D \arrow[r, "s"']
  & E \arrow[r, "t"']
  & F
\end{tikzcd}
\end{center}
\medskip

Left square commutes: $q \circ f = s \circ p$. Right square commutes:
$r \circ g = t \circ q$. The outer rectangle commutes:
$r \circ g \circ f = t \circ s \circ p$. This is the \term{pasting lemma}.

% ─── Simplifying ────────────────────────────────────────────────────────────
\SimplifyingLabel
\subsection{Simplifying: Commutativity in One Line}

A diagram commutes if and only if: for every pair of objects $X$, $Y$ and
every two paths $p_1$, $p_2$ of morphisms from $X$ to $Y$, the composites
are equal.

\[\text{diagram commutes} \iff
  \forall\, X, Y,\; \forall\, \text{paths } p_1, p_2 : X \leadsto Y,\quad
  \mathrm{comp}(p_1) = \mathrm{comp}(p_2)
\]

All of our examples are instances of this single condition.

% ─── Formally ───────────────────────────────────────────────────────────────
\FormallyLabel
\subsection{Formally: The Pasting Lemma}

\formalnote{Pasting lemma}
Let two commutative squares share a common edge. If
$q \circ f = s \circ p$ and $r \circ g = t \circ q$, then
$r \circ g \circ f = t \circ s \circ p$.

\emph{Proof.} $r \circ g \circ f = (r \circ g) \circ f = (t \circ q) \circ f
= t \circ (q \circ f) = t \circ (s \circ p) = (t \circ s) \circ p =
t \circ s \circ p$. \qed

The pasting lemma is used constantly in categorical proofs: large commutative
diagrams are built from smaller confirmed pieces, like tiling a floor. Once
you know it, you can decompose any complex diagram into its square or triangle
sub-diagrams and confirm each in turn.

\formalnote{Diagrams without commutativity}
A diagram that is merely depicted (not asserted to commute) is called a
\term{diagram in $\mathcal{C}$} or a \term{graph of shape $J$ in $\mathcal{C}$}.
The formal definition uses the concept of a functor from an index category $J$,
which we will develop in Chapter~3.


% ═══════════════════════════════════════════════════════════════════════════
\section*{Chapter Summary}
\addcontentsline{toc}{section}{Chapter Summary}
\label{sec:ch1-summary}
% ═══════════════════════════════════════════════════════════════════════════

\begin{summarybox}
\textbf{Objects} are the things in a category. We work with them only through
their connections, never by looking inside.

\textbf{Morphisms} ($f : A \to B$) are the directed connections between
objects. Each has a domain and a codomain.

\textbf{Composition}: if $f : A \to B$ and $g : B \to C$, then
$g \circ f : A \to C$. Composition is associative: $(h \circ g) \circ f =
h \circ (g \circ f)$.

\textbf{Identity}: every object $A$ has a morphism $\mathrm{id}_A : A \to A$
satisfying $f \circ \mathrm{id}_A = f = \mathrm{id}_B \circ f$ for any
$f : A \to B$. It is unique.

\textbf{Commutative diagrams} assert that all paths between the same two
objects compose to the same morphism. They are equations in picture form.

\medskip
\textbf{The reader's three questions}, to be asked of every new example:
\begin{enumerate}
  \item What are the objects?
  \item What are the morphisms (domain, codomain, name)?
  \item Which diagrams commute, and what equations do they assert?
\end{enumerate}
\end{summarybox}

% ═══════════════════════════════════════════════════════════════════════════
\section*{Formal Restatement}
\addcontentsline{toc}{section}{Formal Restatement}
\label{sec:ch1-formal}
% ═══════════════════════════════════════════════════════════════════════════

\begin{formalrestatement}
The following collects the formal content of this chapter. A reader returning
here after completing the chapter should be able to read each item and hear
the informal discussion behind it.

\medskip
\textbf{Morphisms.}\quad
A morphism $f : A \to B$ has domain $A$ and codomain $B$.
The hom-set $\mathrm{Hom}(A, B)$ is the collection of all morphisms from $A$
to $B$. Two morphisms are equal iff they have equal domains, equal codomains,
and are the same arrow.

\medskip
\textbf{Composition.}\quad
There is a family of functions
${\circ : \mathrm{Hom}(B,C) \times \mathrm{Hom}(A,B) \to \mathrm{Hom}(A,C)}$
satisfying associativity:
\[
  h \circ (g \circ f) = (h \circ g) \circ f
\]
for all $f \in \mathrm{Hom}(A,B)$, $g \in \mathrm{Hom}(B,C)$,
$h \in \mathrm{Hom}(C,D)$.

\medskip
\textbf{Identity.}\quad
Each object $A$ carries a distinguished morphism $\mathrm{id}_A \in
\mathrm{Hom}(A,A)$ satisfying the unit laws:
\[
  f \circ \mathrm{id}_A = f = \mathrm{id}_B \circ f
  \qquad \text{for all } f \in \mathrm{Hom}(A, B).
\]
The identity morphism is unique.

\medskip
\textbf{Commutativity.}\quad
A diagram commutes iff for every pair of objects $X$, $Y$ and every two
paths of morphisms from $X$ to $Y$, the two composites are equal.

\medskip
\textbf{Pasting lemma.}\quad
If two commutative squares share a common edge, the outer rectangle commutes.

\medskip
\noindent\textit{Note:} Objects, morphisms, composition, and identity together
form the data of a \term{category} --- to be defined precisely in Chapter~2,
where the axioms above become the definition.
\end{formalrestatement}

\bigskip

In Chapter~2, we assemble all these pieces into a single package: a
\term{category}.
