% ch38_simplicial_sets.tex — Volume IV Chapter 2
% Retrofitted with full Vol I-III pedagogy: cycle stages, etymology, dialogs.

\chapter{Simplicial Sets}

\begin{overviewbox}
With $\Delta$ in hand we can now name the combinatorial structures it
indexes. A \term{simplicial set} is a contravariant functor $X :
\Delta^{\mathrm{op}} \to \mathbf{Set}$ — equivalently, a sequence of
sets $X_0, X_1, X_2, \ldots$ together with face operators $d_i :
X_n \to X_{n-1}$ and degeneracy operators $s_i : X_n \to X_{n+1}$
satisfying the \emph{simplicial identities}, which are the cosimplicial
identities of Chapter~37 with arrows reversed.

This chapter develops the category $\mathbf{sSet} =
\mathbf{Set}^{\Delta^{\mathrm{op}}}$ and its core combinatorial objects:
the \term{standard $n$-simplices} $\Delta^n$, the \term{boundaries}
$\partial\Delta^n$, and the \term{horns} $\Lambda^n_k$. The distinction
between \emph{inner} horns ($0 < k < n$) and \emph{outer} horns ($k = 0$
or $k = n$) is the central technical fact organizing the next chapter
and all of $\infty$-category theory.

Examples drive the chapter. We build the simplicial circle $S^1$ by
gluing two endpoints of an edge, exhibit the nerve $\mathrm{N}(\mathcal{C})$
of an ordinary category (the first taste of Chapter~39), and preview the
singular simplicial set $\mathrm{Sing}(X)$ of a topological space.
Skeleta and coskeleta provide the truncation tools — every simplicial
set is determined by its $n$-simplices for $n$ large enough.
\end{overviewbox}


% ═══════════════════════════════════════════════════════════════════════════
\section{The Category of Simplicial Sets}
\label{sec:sset-defn}
% ═══════════════════════════════════════════════════════════════════════════

\subsection{Informally}

Let us begin with a non-categorical description, before reorganizing
into a clean functor definition.

A \emph{simplicial set} is a recipe for a combinatorial shape built out
of simplices of every dimension. To specify one, we say:
\begin{itemize}
  \item How many $0$-simplices there are (vertices).
  \item How many $1$-simplices there are (edges), each with two endpoints
        (specified by face operators).
  \item How many $2$-simplices there are (triangles), each with three
        edges and three vertices (again specified by face operators).
  \item And so on for each dimension $n$.
\end{itemize}

On top of the counts, we record \emph{how} the simplices fit together:
\begin{itemize}
  \item \term{Face operators} $d_i : X_n \to X_{n-1}$: tell us, for each
        $n$-simplex, what its $(n-1)$-dimensional faces are.
  \item \term{Degeneracy operators} $s_i : X_n \to X_{n+1}$: produce
        higher-dimensional ``degenerate'' simplices by repeating
        vertices.
\end{itemize}

The face and degeneracy operators must satisfy a tower of compatibility
conditions — the \emph{simplicial identities} — that ensure the
combinatorial geometry hangs together coherently. The miracle of
Chapter~37 is that all those compatibility conditions are subsumed by a
single sentence: ``$X$ is a contravariant functor from $\Delta$ to
$\mathbf{Set}$.''

\begin{nameorigin}
\term{Simplicial} is the adjectival form of \emph{simplex} (Latin
\emph{simplex}, ``simple''). A simplicial set is ``a set indexed by
simplices,'' just as a \emph{topological} set is one indexed by topology
or a \emph{measurable} set one indexed by measurable structure. The
terminology dates to Eilenberg–Zilber (1950), who isolated the category
of \emph{semisimplicial} complexes (face maps only); the modern
``simplicial sets'' (with both face and degeneracy operators) was
standardized by Kan in the late 1950s.
\end{nameorigin}

\begin{readernote}
The set of all simplicial-set data is enormous: even small simplicial
sets have infinitely many degeneracies (every simplex of every dimension
generates an infinite tower of higher-dimensional degenerate simplices).
The bookkeeping looks daunting at first. The trick is to focus on the
\emph{non-degenerate} simplices — these are finitely many for many
simplicial sets of interest, and they determine everything else.
\end{readernote}

\subsection{Expanding}

A simplicial set $X$ has \emph{three} layers of structure, building on
one another:

\paragraph{Layer 1: the sets $X_n$.}
For each $n \geq 0$, a set $X_n$ — the ``set of $n$-simplices.''

\paragraph{Layer 2: the face operators.}
For each $i, n$ with $0 \leq i \leq n$, a function $d_i : X_n \to
X_{n-1}$. Reading: ``the $i$-th face operator on $X_n$.'' Applied to an
$n$-simplex $x \in X_n$, $d_i x \in X_{n-1}$ is the face of $x$
opposite vertex $i$.

\paragraph{Layer 3: the degeneracy operators.}
For each $i, n$ with $0 \leq i \leq n$, a function $s_i : X_n \to
X_{n+1}$. Applied to $x \in X_n$, $s_i x$ is the ``degenerate''
$(n+1)$-simplex obtained by repeating vertex $i$.

\paragraph{The simplicial identities.}
The face and degeneracy operators must satisfy a small finite list of
compatibility conditions — the \term{simplicial identities}
(Theorem~\ref{thm:sset-combinatorial}). These are the duals of the
cosimplicial identities of Chapter~37, obtained by reversing arrow order.

\paragraph{Why a functor?}
Specifying the three layers above and verifying the identities is
exactly equivalent to specifying a contravariant functor $X :
\Delta^{\mathrm{op}} \to \mathbf{Set}$, where:
\begin{itemize}
  \item $X([n]) = X_n$.
  \item $X(\delta^i) = d_i$.
  \item $X(\sigma^i) = s_i$.
\end{itemize}
The simplicial identities become automatic consequences of $X$ being a
functor on $\Delta^{\mathrm{op}}$ (functoriality $=$ all relations among
generators preserved). This functorial repackaging is the modern style.

\begin{dialog}
\textbf{Reader:} Why a \emph{contravariant} functor? Why not just go in
the same direction as $\Delta$?

\textbf{Author:} The contravariance lines up the geometric expectation
with the algebra. In $\Delta$, the face map $\delta^i : [n-1] \to [n]$
goes \emph{up} in dimension. A simplicial set $X$ assigns sets to
ordinals, and a face operator $d_i : X_n \to X_{n-1}$ goes \emph{down}
in dimension (because a face has \emph{lower} dimension than its
ambient simplex). To get the directions to match — face maps go up,
face operators go down — we use the opposite category.

\textbf{Reader:} So \emph{cosimplicial} objects, which use $\Delta$
directly, have the dimensions going up?

\textbf{Author:} Yes. Cosimplicial = covariant on $\Delta$ = ``dimensions
go up.'' Simplicial = contravariant on $\Delta$ = ``dimensions go down.''
We will see both kinds: cosimplicial spaces (Chapter~37); simplicial
spaces (this chapter).
\end{dialog}

\begin{nameorigin}
A \term{presheaf} on a category $\mathcal{C}$ is a contravariant functor
$\mathcal{C}^{\mathrm{op}} \to \mathbf{Set}$. The name \emph{presheaf}
predates the modern definition: Jean Leray (1946) introduced
\emph{faisceau} (French ``sheaf''), and \emph{préfaisceau} (presheaf)
was coined for the underlying data of a sheaf \emph{before} imposing
the gluing axiom. ``Pre-'' from Latin \emph{prae-} (``before, in front
of'') — the data exists \emph{prior to} the sheaf condition. Simplicial
sets are presheaves on $\Delta$.
\end{nameorigin}

\subsection{Formalizing}

\begin{definition}[Simplicial set]
\label{def:simplicial-set}
A \term{simplicial set} is a functor
\begin{equation*}
  X : \Delta^{\mathrm{op}} \to \mathbf{Set}.
\end{equation*}
We write $X_n = X([n])$ and call its elements \term{$n$-simplices} of $X$.
For each $\delta^i : [n-1] \to [n]$ in $\Delta$, the contravariant
$X(\delta^i)$ is a function $X_n \to X_{n-1}$ written $d_i$; for each
$\sigma^i : [n+1] \to [n]$, the contravariant $X(\sigma^i)$ is a function
$X_n \to X_{n+1}$ written $s_i$.

A \term{morphism} of simplicial sets is a natural transformation
$X \to Y$. The category of simplicial sets is denoted $\mathbf{sSet} :=
[\Delta^{\mathrm{op}}, \mathbf{Set}]$.
\end{definition}

\begin{howtosay}
$\mathbf{sSet}$ \sayit{``simplicial sets''}; $X_n$ \sayit{``the $n$-simplices
of $X$''} or \sayit{``$X$ at level $n$''}; $d_i$ \sayit{``$i$-th face of''} or
\sayit{``$d$-sub-$i$''}; $s_i$ \sayit{``$i$-th degeneracy of''} or \sayit{``$s$-sub-$i$''}.
The boldface $\mathbf{sSet}$ distinguishes the category from individual
simplicial sets; some texts write $\mathrm{Set}_\Delta$ or
$\widehat{\Delta}$ (the latter using the standard ``hat'' notation for
presheaf categories).
\end{howtosay}

\begin{readernote}
The notation $d_i$ and $s_i$ uses subscripts where Chapter~37 used
superscripts ($\delta^i$, $\sigma^i$). This is intentional: the
superscript variants live in $\Delta$ itself (covariant); the subscript
variants are induced by the contravariant functor $X$, so they ``go the
other way.'' Keep the subscript/superscript convention in mind whenever
you see a face or degeneracy.
\end{readernote}

\begin{theorem}[Equivalent combinatorial description]
\label{thm:sset-combinatorial}
The data of a simplicial set $X : \Delta^{\mathrm{op}} \to \mathbf{Set}$ is
equivalent to the data of:
\begin{itemize}
  \item A sequence of sets $\{X_n\}_{n \geq 0}$.
  \item For each $0 \leq i \leq n$, a face operator $d_i : X_n \to X_{n-1}$.
  \item For each $0 \leq i \leq n$, a degeneracy operator $s_i : X_n \to X_{n+1}$.
\end{itemize}
subject to the \term{simplicial identities}:
\begin{align*}
  d_i d_j      &= d_{j-1} d_i      & & \text{if } i < j, \\
  s_i s_j      &= s_{j+1} s_i      & & \text{if } i \leq j, \\
  d_i s_j      &= s_{j-1} d_i      & & \text{if } i < j, \\
  d_i s_j      &= \mathrm{id}      & & \text{if } i = j \text{ or } i = j+1, \\
  d_i s_j      &= s_j d_{i-1}      & & \text{if } i > j+1.
\end{align*}
\end{theorem}

\begin{proof}[Proof sketch]
The face/degeneracy maps $\delta^i, \sigma^i$ in $\Delta$ generate all
morphisms (Theorem~37.\ref{thm:delta-generators}). Hence specifying a
functor $X$ on $\Delta^{\mathrm{op}}$ is the same as specifying its values
on $\delta^i$ and $\sigma^i$, namely the $d_i$ and $s_i$. The relations
that hold between $\delta^i$ and $\sigma^i$ (the cosimplicial identities)
become, under the contravariant $X$, the simplicial identities above.
\qed
\end{proof}

\begin{remark}[Dualizing the indexing]
The simplicial identities differ from the cosimplicial identities of
Chapter~37 only by reversed arrow order: $\delta^j \delta^i$ in $\Delta$
becomes $d_i d_j$ in $\Delta^{\mathrm{op}}$ (the order of composition flips
under the duality). The shift constants $\pm 1$ also flip sides; this is
the source of the off-by-one variations between cosimplicial and simplicial
identities that one sees in different references.
\end{remark}


% ═══════════════════════════════════════════════════════════════════════════
\section{Standard Simplices, Boundaries, and Horns}
\label{sec:simplices-boundaries-horns}
% ═══════════════════════════════════════════════════════════════════════════

\subsection{Informally}

There are three foundational simplicial sets that every other will be
built from:
\begin{itemize}
  \item The \term{standard $n$-simplex} $\Delta^n$: the ``free'' simplicial
        set with one distinguished non-degenerate $n$-simplex. This is
        the Yoneda image of $[n] \in \Delta$.
  \item The \term{boundary} $\partial\Delta^n$: the standard $n$-simplex
        with its top-dimensional simplex removed, leaving only the
        ``shell.''
  \item The \term{horn} $\Lambda^n_k$: the boundary with one more face
        ($\delta^k$) also removed, leaving an open ``$U$-shaped'' or
        ``$\mathsf{L}$-shaped'' partial simplex.
\end{itemize}
These three sit in a sequence
$\Lambda^n_k \;\hookrightarrow\; \partial\Delta^n \;\hookrightarrow\; \Delta^n$,
removing one $n$-simplex and one face at a time. The combinatorics of
filling horns and boundaries — extending a partial map back up to a
full $\Delta^n$-map — is the central technical content of
$\infty$-category theory.

\subsection{Formalizing}

\begin{definition}[Standard $n$-simplex]
\label{def:standard-simplex}
The \term{standard $n$-simplex} is the simplicial set $\Delta^n$
representable by $[n] \in \Delta$:
\begin{equation*}
  \Delta^n \;=\; \Delta(-, [n]) : \Delta^{\mathrm{op}} \to \mathbf{Set}.
\end{equation*}
Its $k$-simplices are the monotone maps $[k] \to [n]$. The face operators
act by precomposition with $\delta^i : [k-1] \to [k]$, and the degeneracy
operators by precomposition with $\sigma^i : [k+1] \to [k]$.
\end{definition}

\begin{howtosay}
$\Delta^n$ \sayit{``Delta-$n$'' or ``the standard $n$-simplex''}; the
``standard'' modifier distinguishes the simplicial set $\Delta^n$ from
the topological simplex $\lvert\Delta^n\rvert$
(\S37.\ref{sec:geometric-realization-preview}). Context usually makes
which is meant clear; when ambiguous, ``combinatorial'' vs.\ ``topological''
disambiguates.
\end{howtosay}

\begin{nameorigin}
The notation $\Delta^n$ (with superscript $n$, not the bracket $[n]$ of
$\Delta$ itself) names the \emph{simplicial set represented by} $[n]$
under the Yoneda embedding $\Delta \hookrightarrow \mathbf{sSet}$. The
convention comes from algebraic topology, where the ``standard
$n$-simplex'' was the canonical name for the geometric object
$\lvert\Delta^n\rvert$; when the combinatorial avatar emerged, the
notation $\Delta^n$ (without bars) was kept for the simplicial set.
\end{nameorigin}

\begin{example_thm}[Low-dimensional $\Delta^n$]
$\Delta^0$ is the constant simplicial set with one $k$-simplex in every
dimension (the unique monotone $[k] \to [0]$). It is the terminal object
of $\mathbf{sSet}$. $\Delta^1$ has two $0$-simplices (the maps $[0] \to
[1]$ named $0$ and $1$), three $1$-simplices (the identity and the two
constants), and a growing number of degenerate $k$-simplices for $k \geq 2$.
\end{example_thm}

\begin{lemma}[Yoneda for simplicial sets]
\label{lem:yoneda-sset}
For any simplicial set $X$, the set of morphisms $\Delta^n \to X$ in
$\mathbf{sSet}$ is naturally bijective with $X_n$:
\begin{equation*}
  \mathbf{sSet}(\Delta^n, X) \;\cong\; X_n.
\end{equation*}
\end{lemma}

\begin{proof}
This is the Yoneda lemma (Chapter~9 / Chapter~15) applied to the
representable $\Delta^n = \Delta(-, [n])$ in $\mathbf{sSet} =
[\Delta^{\mathrm{op}}, \mathbf{Set}]$. The bijection sends $f : \Delta^n
\to X$ to $f_{[n]}(\mathrm{id}_{[n]}) \in X_n$.
\qed
\end{proof}

\begin{readernote}
Yoneda for $\mathbf{sSet}$ is the single most-used computational fact
in the chapter. It says: \emph{``an $n$-simplex of $X$ is the same as a
map $\Delta^n \to X$''}. We will switch fluidly between the two
descriptions. A given $n$-simplex $x \in X_n$ corresponds to its
``picking-out map'' $\widehat{x} : \Delta^n \to X$, and the universal
$n$-simplex of $\Delta^n$ itself is $\mathrm{id}_{[n]} \in \Delta^n_n$.
\end{readernote}

\begin{definition}[Boundary of $\Delta^n$]
\label{def:boundary}
The \term{boundary} $\partial\Delta^n \subset \Delta^n$ is the simplicial
subset whose $k$-simplices are the \emph{non-surjective} monotone maps
$[k] \to [n]$. Equivalently, $\partial\Delta^n$ is the union of the
images of the $n+1$ face maps $\delta^i : \Delta^{n-1} \to \Delta^n$,
\begin{equation*}
  \partial\Delta^n \;=\; \bigcup_{i=0}^{n} \delta^i(\Delta^{n-1}).
\end{equation*}
For $n = 0$, we set $\partial\Delta^0 = \emptyset$ (the empty simplicial set).
\end{definition}

\begin{nameorigin}
The symbol \term{$\partial$} for ``boundary'' is from classical topology,
where it denoted the boundary operator on chain complexes (Poincaré's
work in the late 1800s). The combinatorial boundary $\partial\Delta^n$ is
the simplicial set version: it removes the one top-dimensional
non-degenerate simplex, leaving the union of the $n+1$ codimension-$1$
faces. Geometrically, $\lvert\partial\Delta^n\rvert$ is homeomorphic to
the $(n-1)$-sphere $S^{n-1}$.
\end{nameorigin}

\begin{definition}[Horn $\Lambda^n_k$]
\label{def:horn}
For $0 \leq k \leq n$ with $n \geq 1$, the \term{$k$-th horn} $\Lambda^n_k$
is the simplicial subset of $\partial\Delta^n$ obtained by omitting the
$k$-th face:
\begin{equation*}
  \Lambda^n_k \;=\; \bigcup_{i \neq k} \delta^i(\Delta^{n-1}).
\end{equation*}
We call $\Lambda^n_k$ an \term{inner horn} if $0 < k < n$, and an
\term{outer horn} if $k = 0$ or $k = n$.
\end{definition}

\begin{howtosay}
$\Lambda^n_k$ \sayit{``Lambda $n$-$k$'' or ``the $k$-th horn of $\Delta^n$''}.
The geometric picture: $\Delta^n$ with the interior \emph{and} the face
opposite vertex $k$ removed, leaving an open ``horn'' that bends toward
the missing $k$-th face.
\end{howtosay}

\begin{nameorigin}
The letter \term{$\Lambda$} (Greek capital lambda, ``L'') was chosen by
Joyal in the late 1990s for the horn simplicial set because:
\begin{itemize}
  \item The shape $\Lambda$ visually resembles the case $n = 2$:
        $\Lambda^2_1$ is two edges joined at the middle vertex,
        forming a wide ``$\Lambda$'' shape (or an inverted ``$\mathsf{V}$'').
  \item More mundanely, $\Lambda$ is one of the few unused Greek
        capitals close to $\Delta$, making the typographic pairing
        natural.
\end{itemize}
The terminology \term{horn} reflects the geometric intuition: $\Lambda^n_k$
is the boundary $\partial\Delta^n$ with one face removed, leaving an
``open horn'' that bends or curves around the missing face.
\end{nameorigin}

\begin{example_thm}[Horns of low dimension]
\label{ex:horns-low}
$\Lambda^2_1$ is the inner horn of the $2$-simplex (triangle): two edges
$0 \to 1$ and $1 \to 2$ joined at the middle vertex, missing the long
edge $0 \to 2$ and the $2$-cell. $\Lambda^2_0$ and $\Lambda^2_2$ are the
two outer horns, each consisting of two edges joined at an endpoint
(forming an ``$\mathsf{L}$''-shape).

$\Lambda^3_1$ and $\Lambda^3_2$ are the two inner horns of the
$3$-simplex (tetrahedron): three of the four triangular faces, missing
either the face opposite vertex $1$ or the face opposite vertex $2$.
\end{example_thm}

\subsection{Reorganizing: Inner vs.\ outer horns}

The inner/outer horn distinction is the central technical fact of
$\infty$-category theory.

\begin{nameorigin}
The distinction between \term{inner} and \term{outer} horns refers to
\emph{which vertex} is at the bend:
\begin{itemize}
  \item \emph{Inner} ($0 < k < n$): the missing face is opposite an
        ``interior'' vertex (not the first or last). The horn closes
        around an inner vertex.
  \item \emph{Outer} ($k = 0$ or $k = n$): the missing face is opposite
        the first or last vertex. The horn is at the ``outer'' edge of
        the index range.
\end{itemize}
The terminology dates to Boardman–Vogt (1973), who originally called
the inner-horn-filling condition the ``restricted Kan condition.''
Joyal's preferred ``inner/outer'' terminology stuck.
\end{nameorigin}

\begin{remark}[Foreshadowing the Kan condition]
A simplicial set in which every inner horn $\Lambda^n_k \to X$ (with
$0 < k < n$) extends to $\Delta^n \to X$ is called a \term{quasi-category}
(Chapter~43). If every horn (inner or outer) extends, it is called a
\term{Kan complex} (Chapter~39).

The composition law of a quasi-category arises from extending inner
$2$-horns: a $\Lambda^2_1 \to X$ is a pair of composable $1$-simplices,
and the existence of a $\Delta^2$ filling provides their composite as the
long edge $0 \to 2$. The inner-horn condition makes composition
existential but non-unique; the choice of filler is the witness that an
$\infty$-category has \emph{coherent} rather than strict associativity.
\end{remark}

\begin{readernote}
The mnemonic: \emph{inner horns let you compose; outer horns let you
invert}. Filling $\Lambda^2_1$ (inner) takes two composable arrows and
produces their composite. Filling $\Lambda^2_0$ (outer) takes
$f, g : a \to b$ with $f$ given, and asks for an arrow $h : a \to a$
making the triangle commute — effectively asking $f$ to have a
``left inverse.'' Quasi-categories impose only the composition
condition; Kan complexes (= $\infty$-groupoids) impose both.
\end{readernote}

\subsection{Formally}

\begin{lemma}[$k$-simplices of $\partial\Delta^n$ and $\Lambda^n_k$]
\label{lem:bdry-horn-simplices}
The $k$-simplices of $\partial\Delta^n$ are the monotone maps $[k] \to [n]$
that miss at least one element; the $k$-simplices of $\Lambda^n_j$ are
those missing at least one element other than $j$.
\end{lemma}


% ═══════════════════════════════════════════════════════════════════════════
\section{Skeleta and Coskeleta}
\label{sec:skeleta}
% ═══════════════════════════════════════════════════════════════════════════

\subsection{Informally}

The category $\mathbf{sSet}$ has objects of \emph{every} dimension, so
it is reasonable to ask: \emph{how much information is in the first $n$
dimensions of a simplicial set?} Two limit-style answers:

\begin{itemize}
  \item Take the simplicial subset generated by the $\leq n$-simplices,
        with higher dimensions added back only as degeneracies of those.
        This is the \term{$n$-skeleton} $\mathrm{sk}_n X$.
  \item Take the largest simplicial set whose $\leq n$-simplices agree
        with $X$'s, freely adding back simplices of higher dimension
        subject only to boundary constraints. This is the \term{$n$-coskeleton}
        $\mathrm{cosk}_n X$.
\end{itemize}

Both come from the same kind of construction (Kan extension along a
truncation functor), but they go in opposite directions: $\mathrm{sk}_n$
adds as little as possible, $\mathrm{cosk}_n$ adds as much as possible.

\begin{nameorigin}
The terms \term{skeleton} and \term{coskeleton} are loans from
algebraic topology and the theory of CW complexes. A CW complex's
\emph{$n$-skeleton} is the union of its cells of dimension $\leq n$ — the
skeletal structure beneath the higher cells. In simplicial sets, the
$n$-skeleton plays the same role: the part of $X$ in dimensions
$\leq n$, with everything higher derived from it.

The prefix \emph{co-} in ``coskeleton'' signals \emph{dual} (the right
adjoint instead of the left), matching the long category-theoretic
tradition: cofibration, colimit, cohomology — all duals of the
corresponding ``un-prefixed'' notions.
\end{nameorigin}

\subsection{Formalizing}

For each $n \geq 0$ there is a full subcategory $\Delta_{\leq n} \subset
\Delta$ on objects $[0], [1], \ldots, [n]$. The corresponding restriction
functor on presheaves yields a chain of adjoints that decomposes
$\mathbf{sSet}$ by dimension.

\begin{definition}[$n$-skeleton, $n$-coskeleton]
\label{def:skel-cosk}
Let $\iota_n : \Delta_{\leq n} \hookrightarrow \Delta$ be the inclusion.
The restriction functor
\begin{equation*}
  \iota_n^* : \mathbf{sSet} \longrightarrow \mathbf{Set}^{\Delta_{\leq n}^{\mathrm{op}}}
\end{equation*}
has both a left adjoint $\mathrm{sk}_n$ (left Kan extension, the
\term{$n$-skeleton}) and a right adjoint $\mathrm{cosk}_n$ (right Kan
extension, the \term{$n$-coskeleton}):
\begin{equation*}
  \mathrm{sk}_n \;\dashv\; \iota_n^* \;\dashv\; \mathrm{cosk}_n.
\end{equation*}
A simplicial set $X$ is \term{$n$-skeletal} if the counit
$\mathrm{sk}_n \iota_n^* X \to X$ is an isomorphism; it is
\term{$n$-coskeletal} if the unit $X \to \mathrm{cosk}_n \iota_n^* X$ is.
\end{definition}

\begin{howtosay}
$\mathrm{sk}_n X$ \sayit{``the $n$-skeleton of $X$''}; $\mathrm{cosk}_n X$
\sayit{``the $n$-coskeleton of $X$''}. ``Skeletal'' and ``coskeletal''
are adjectives applied to simplicial sets: $X$ is \emph{$n$-skeletal}
if it agrees with its own $n$-skeleton; $X$ is \emph{$n$-coskeletal} if
it agrees with its own $n$-coskeleton.
\end{howtosay}

\begin{remark}[The skeleta in pictures]
$\mathrm{sk}_n X$ is the smallest simplicial subset of $X$ containing
all $k$-simplices of $X$ for $k \leq n$, with all simplices of higher
dimension being degeneracies of those. $\mathrm{cosk}_n X$ is the largest
simplicial set agreeing with $X$ in dimensions $\leq n$: it freely adds
back $k$-simplices for $k > n$ subject only to the boundary constraints.

In particular: $\Delta^n$ is $n$-skeletal (it is generated by its unique
non-degenerate $n$-simplex $\mathrm{id}_{[n]}$); $\partial\Delta^n$ is
$(n-1)$-skeletal (it has no non-degenerate $n$-simplex); $\Lambda^n_k$ is
also $(n-1)$-skeletal.
\end{remark}

\begin{lemma}[Skeleton filtration]
\label{lem:skeleton-filtration}
For any $X \in \mathbf{sSet}$, the canonical maps assemble into a filtration
\begin{equation*}
  \emptyset \;=\; \mathrm{sk}_{-1} X \;\hookrightarrow\; \mathrm{sk}_0 X \;\hookrightarrow\; \mathrm{sk}_1 X \;\hookrightarrow\; \cdots \;\hookrightarrow\; X,
\end{equation*}
with $X = \mathrm{colim}_n \mathrm{sk}_n X$.
\end{lemma}

\begin{readernote}
The skeleton filtration is the simplicial-set version of the cell-complex
filtration in topology. Inductive arguments on simplicial sets routinely
proceed up this filtration: prove the statement for $\mathrm{sk}_n X$,
assuming it for $\mathrm{sk}_{n-1} X$. The step usually amounts to
filling in non-degenerate $n$-simplices one at a time, via boundary
inclusions $\partial\Delta^n \hookrightarrow \Delta^n$. We will see this
argument pattern repeatedly in Chapters~41 (model categories) and 43
(quasi-categories).
\end{readernote}


% ═══════════════════════════════════════════════════════════════════════════
\section{Examples}
\label{sec:sset-examples}
% ═══════════════════════════════════════════════════════════════════════════

\subsection{The simplicial circle}

\begin{example_thm}[The simplicial circle $S^1$]
\label{ex:simplicial-circle}
Let $S^1 = \Delta^1 / \partial\Delta^1$ — the simplicial set obtained from
$\Delta^1$ by identifying the two endpoints. In $S^1$ there is one
non-degenerate $0$-simplex $*$ (the common image of the endpoints) and one
non-degenerate $1$-simplex $\ell$ (the image of the edge of $\Delta^1$),
both face operators of which return $*$. All higher-dimensional
non-degenerate simplices are absent; the higher $k$-simplices are
exclusively degeneracies $s_{i_1} \cdots s_{i_p} \ell$ or $s_{i_1} \cdots
s_{i_p} *$.

Geometrically, $S^1$'s realization (Chapter~39) is the topological circle.
\end{example_thm}

\subsection{The nerve of a category (preview)}

\begin{nameorigin}
The \term{nerve} of a category was introduced by Alexander Grothendieck
in correspondence around 1960 and developed in detail by Graeme Segal
(1968) in his work on classifying spaces of categories and topological
operads. The word ``nerve'' is a biological metaphor: just as the nervous
system of an organism is the structural backbone that transmits
information throughout the body, the nerve of a category is the
simplicial structure that ``transmits'' the categorical information into
the world of simplicial sets. The metaphor predates Grothendieck — Hilbert
used ``nerve of a covering'' (Nerv einer Bedeckung) in topology around
1932 — but Grothendieck's categorical nerve is the modern usage.
\end{nameorigin}

\begin{example_thm}[Nerve of a category]
\label{ex:nerve-preview}
For an ordinary category $\mathcal{C}$, the \term{nerve} $\mathrm{N}(\mathcal{C}) \in
\mathbf{sSet}$ has:
\begin{itemize}
  \item $\mathrm{N}(\mathcal{C})_0 = \mathrm{Ob}(\mathcal{C})$;
  \item $\mathrm{N}(\mathcal{C})_1 = \mathrm{Mor}(\mathcal{C})$;
  \item $\mathrm{N}(\mathcal{C})_n = \{\,c_0 \xrightarrow{f_1} c_1 \xrightarrow{f_2} \cdots \xrightarrow{f_n} c_n : n\text{-fold composable chain}\,\}$.
\end{itemize}
Face operators compose or drop end morphisms; degeneracy operators insert
identities.
\end{example_thm}

\begin{remark}[A preview of the next chapter]
The nerve construction defines a functor $\mathrm{N} : \mathbf{Cat} \to
\mathbf{sSet}$ that is fully faithful (Chapter~39) and whose image consists
of exactly those simplicial sets in which every inner horn extends
\emph{uniquely}. That uniqueness — versus the merely-existence-of-a-filler
of quasi-categories — is what distinguishes ordinary $1$-categories from
$\infty$-categories within $\mathbf{sSet}$.
\end{remark}

\subsection{The singular complex of a space (preview)}

\begin{nameorigin}
The \term{singular} simplicial set, \term{$\mathrm{Sing}(X)$}, was
introduced by Samuel Eilenberg (1944) as the foundation of singular
homology. The adjective ``singular'' contrasts with ``smooth'' or
``polyhedral'': a singular simplex in $X$ is merely a continuous map
$\lvert\Delta^n\rvert \to X$, which can land on or pass through
``singular points'' (corners, edges, non-smooth locations) of $X$.
Earlier theories required smoother conditions; Eilenberg's insight was
that allowing arbitrary continuous maps gave a homology theory defined
for \emph{all} topological spaces.
\end{nameorigin}

\begin{example_thm}[Singular complex $\mathrm{Sing}(X)$]
\label{ex:singular}
For a topological space $X$, the \term{singular simplicial set} is
\begin{equation*}
  \mathrm{Sing}(X)_n \;=\; \mathbf{Top}\bigl(\lvert\Delta^n\rvert, X\bigr),
\end{equation*}
the set of continuous maps from the standard topological $n$-simplex to
$X$. Face and degeneracy operators come from precomposition with the
cosimplicial space $\lvert\Delta^\bullet\rvert$ of Chapter~37.
\end{example_thm}

\begin{remark}[Sing is a Kan complex]
For any space $X$, every horn $\Lambda^n_k \to \mathrm{Sing}(X)$ extends
to $\Delta^n \to \mathrm{Sing}(X)$ — both inner and outer horns. This is
the classical \term{Kan condition}, and the proof uses only the
contractibility of $\lvert\Lambda^n_k\rvert$ inside $\lvert\Delta^n\rvert$.
Kan complexes will be our model for ``spaces up to homotopy'' in
Chapter~39.
\end{remark}

\begin{readernote}
The pair (nerve, singular complex) gives two contrasting examples of
simplicial sets:
\begin{itemize}
  \item Nerve of a category: \emph{rigid}; inner horns extend uniquely;
        outer horns might not extend at all.
  \item Singular complex of a space: \emph{flexible}; \emph{all} horns
        extend; the extensions are non-unique (every continuous
        deformation gives one).
\end{itemize}
The two examples bookend the spectrum: Chapter~39 develops both, and
Chapter~43 introduces quasi-categories as the in-between case
(inner horns extend, but possibly not uniquely).
\end{readernote}


% ═══════════════════════════════════════════════════════════════════════════
\section{Exercises}
\label{sec:sset-exercises}
% ═══════════════════════════════════════════════════════════════════════════

\begin{exercisesbox}
\begin{qa}
\qaitem{What is a simplicial set?}{A functor $X : \Delta^{\mathrm{op}} \to \mathbf{Set}$; equivalently, a sequence of sets $X_n$ with face and degeneracy operators satisfying the simplicial identities.}
\qaitem{What is $X_n$?}{The set of $n$-simplices of $X$, i.e., $X([n])$.}
\qaitem{Why is the indexing $\Delta^{\mathrm{op}}$ and not $\Delta$?}{So that face operators decrease dimension: $d_i : X_n \to X_{n-1}$. A covariant functor $\Delta \to \mathbf{Set}$ is a \emph{cosimplicial} set, where ``face'' goes the other way.}
\qaitem{Where does ``presheaf'' come from?}{Leray (1946) coined \emph{préfaisceau} as the data of a sheaf \emph{before} (Latin \emph{prae-}) imposing the gluing axiom. A simplicial set is a presheaf on $\Delta$.}
\qaitem{What are the face operators $d_i$?}{For each $0 \leq i \leq n$, $d_i = X(\delta^i) : X_n \to X_{n-1}$. They project an $n$-simplex onto its $i$-th $(n-1)$-face.}
\qaitem{Why subscripts $d_i$, $s_i$ rather than superscripts?}{To signal direction: $\delta^i, \sigma^i$ live in $\Delta$ (covariant); $d_i, s_i$ are induced by the contravariant functor $X$ and thus ``go the other way.''}
\qaitem{State one simplicial identity.}{$d_i s_i = \mathrm{id}$ (and $d_{i+1} s_i = \mathrm{id}$): degenerate then face-extract returns the original.}
\qaitem{What is the standard $n$-simplex $\Delta^n$?}{The representable simplicial set $\Delta(-, [n])$. Its $k$-simplices are the monotone maps $[k] \to [n]$.}
\qaitem{State Yoneda for simplicial sets.}{$\mathbf{sSet}(\Delta^n, X) \cong X_n$, natural in both variables. An $n$-simplex of $X$ is a map $\Delta^n \to X$.}
\qaitem{What is the boundary $\partial\Delta^n$?}{The simplicial subset of $\Delta^n$ consisting of non-surjective monotone maps into $[n]$. Equivalently, the union of the $n+1$ face inclusions $\Delta^{n-1} \hookrightarrow \Delta^n$.}
\qaitem{Why the symbol $\partial$?}{From topology: $\partial$ is the classical boundary operator on chains (Poincaré, late 1800s). The simplicial $\partial\Delta^n$ is the combinatorial version of the topological boundary sphere $S^{n-1}$.}
\qaitem{What is the horn $\Lambda^n_k$?}{The simplicial subset of $\partial\Delta^n$ obtained by removing the $k$-th face: $\Lambda^n_k = \bigcup_{i \neq k} \delta^i(\Delta^{n-1})$.}
\qaitem{Why the letter $\Lambda$ for horns?}{Joyal's choice. The Greek capital lambda visually resembles $\Lambda^2_1$ — two edges joined at a middle vertex form an inverted ``V'' shape, like the letter $\Lambda$. The term \emph{horn} reflects the geometric ``open'' shape with one face removed.}
\qaitem{Inner vs.\ outer horns?}{Inner: $0 < k < n$ (the missing face is opposite an interior vertex). Outer: $k = 0$ or $k = n$ (opposite the first or last vertex). The distinction is central to the Kan/quasi-category dichotomy.}
\qaitem{Picture $\Lambda^2_1$ and $\Lambda^2_0$.}{$\Lambda^2_1$: two edges $0 \to 1$ and $1 \to 2$ joined at the middle (inner). $\Lambda^2_0$: two edges sharing endpoint $0$ (outer, ``$\mathsf{L}$''-shape).}
\qaitem{Inner horn vs.\ outer horn mnemonic?}{\emph{Inner horns let you compose; outer horns let you invert.} Filling $\Lambda^2_1$ produces a composite; filling $\Lambda^2_0$ produces a left inverse.}
\qaitem{When does a horn fill?}{Inner horn fills always: quasi-categories. All horns fill: Kan complexes.}
\qaitem{What is the $n$-skeleton $\mathrm{sk}_n X$?}{The smallest simplicial subset of $X$ generated by simplices of dimension $\leq n$. Equivalently: the left Kan extension along $\Delta_{\leq n} \hookrightarrow \Delta$ applied to $X|_{\Delta_{\leq n}}$.}
\qaitem{What is the $n$-coskeleton $\mathrm{cosk}_n X$?}{The right Kan extension: the largest simplicial set agreeing with $X$ in dimensions $\leq n$. Higher simplices are freely filled subject to boundary constraints.}
\qaitem{Where does ``skeleton'' come from?}{From CW-complex topology: the $n$-skeleton of a CW complex is the union of its cells of dimension $\leq n$ — the structural backbone beneath higher cells. The simplicial-set version plays the same role.}
\qaitem{Why are $\Delta^n, \partial\Delta^n, \Lambda^n_k$ all $n$- or $(n-1)$-skeletal?}{Because they are built from finitely many low-dimensional simplices and their degenerations: $\Delta^n$ is generated by one non-degenerate $n$-simplex (so $n$-skeletal); $\partial\Delta^n$ and $\Lambda^n_k$ have no non-degenerate $n$-simplex (so $(n-1)$-skeletal).}
\qaitem{Describe the simplicial circle $S^1$.}{$S^1 = \Delta^1 / \partial\Delta^1$. One non-degenerate $0$-simplex and one non-degenerate $1$-simplex, both face operators of the latter returning the former.}
\qaitem{What are the $n$-simplices of $\mathrm{N}(\mathcal{C})$?}{$n$-fold composable chains $c_0 \to c_1 \to \cdots \to c_n$ of morphisms in $\mathcal{C}$.}
\qaitem{Where does ``nerve'' come from?}{Biological metaphor (Grothendieck $\sim$ 1960; Segal 1968). The nervous system transmits information through an organism; the nerve of a category transmits its categorical information into $\mathbf{sSet}$.}
\qaitem{Faces of an $n$-simplex in the nerve?}{$d_0$ drops the leftmost morphism (and object); $d_n$ drops the rightmost; for $0 < i < n$, $d_i$ composes adjacent morphisms $f_{i+1} \circ f_i$ to shorten the chain.}
\qaitem{What is the singular simplicial set $\mathrm{Sing}(X)$?}{$\mathrm{Sing}(X)_n = \mathbf{Top}(\lvert\Delta^n\rvert, X)$: continuous maps from the topological $n$-simplex into $X$.}
\qaitem{Where does ``singular'' come from?}{Eilenberg (1944): allowed arbitrary continuous maps from $\lvert\Delta^n\rvert$ into $X$, not just smooth or polyhedral ones. ``Singular'' contrasts with ``smooth'' — the simplex may pass through singular points of $X$.}
\qaitem{Is $\mathrm{Sing}(X)$ a Kan complex?}{Yes — every horn $\Lambda^n_k \to \mathrm{Sing}(X)$ extends to $\Delta^n \to \mathrm{Sing}(X)$, because $\lvert\Lambda^n_k\rvert$ is a deformation retract of $\lvert\Delta^n\rvert$.}
\qaitem{Why does Yoneda for $\mathbf{sSet}$ pay off practically?}{It identifies maps out of $\Delta^n$ with $n$-simplices of the target — converting universal mapping properties into concrete simplex-counting statements.}
\qaitem{What is $\mathrm{sk}_0 X$?}{The discrete simplicial set on $X_0$: just the $0$-simplices of $X$, with all higher simplices degenerate.}
\qaitem{Is $\mathbf{sSet}$ complete and cocomplete?}{Yes — as a presheaf category $[\Delta^{\mathrm{op}}, \mathbf{Set}]$, it inherits all (co)limits from $\mathbf{Set}$, computed pointwise (Chapter~6).}
\end{qa}
\end{exercisesbox}


% ═══════════════════════════════════════════════════════════════════════════
\section*{Chapter Summary}
\addcontentsline{toc}{section}{Chapter Summary}
% ═══════════════════════════════════════════════════════════════════════════

\begin{summarybox}
\textbf{Simplicial sets.}\quad $\mathbf{sSet} = [\Delta^{\mathrm{op}},
\mathbf{Set}]$. A simplicial set is a sequence of sets $X_n$ together
with face operators $d_i$ and degeneracy operators $s_i$ satisfying the
simplicial identities (the dual of the cosimplicial identities of
Chapter~37).

\textbf{Standard simplices, boundaries, horns.}\quad $\Delta^n$ is the
representable simplicial set; an $n$-simplex of $X$ is a map $\Delta^n
\to X$ (Yoneda). $\partial\Delta^n$ is the boundary (non-surjective
monotone maps into $[n]$); $\Lambda^n_k$ is the horn omitting the
$k$-th face. \emph{Inner} horns have $0 < k < n$; \emph{outer} horns
have $k = 0$ or $k = n$. Mnemonic: \emph{inner horns let you compose;
outer horns let you invert}.

\textbf{Skeleta and coskeleta.}\quad $\mathrm{sk}_n \dashv \iota_n^*
\dashv \mathrm{cosk}_n$ gives the canonical filtration $\emptyset \subset
\mathrm{sk}_0 X \subset \mathrm{sk}_1 X \subset \cdots$, with $X$ as the
colimit. $\Delta^n$ is $n$-skeletal; $\partial\Delta^n$ and $\Lambda^n_k$
are $(n-1)$-skeletal. Borrowed from CW-complex topology.

\textbf{Key examples.}\quad The simplicial circle $S^1 = \Delta^1 /
\partial\Delta^1$; the nerve $\mathrm{N}(\mathcal{C})$ of a category
(Grothendieck/Segal, $\sim$ 1968); the singular simplicial set
$\mathrm{Sing}(X)$ of a space (Eilenberg, 1944). The latter is a Kan
complex — all horns fill, not merely the inner ones. The distinction
foreshadows quasi-categories (Chapter~43).
\end{summarybox}


% ═══════════════════════════════════════════════════════════════════════════
\section*{Formal Restatement}
\addcontentsline{toc}{section}{Formal Restatement}
% ═══════════════════════════════════════════════════════════════════════════

\begin{formalrestatement}
\textbf{Category.}\quad $\mathbf{sSet} = [\Delta^{\mathrm{op}}, \mathbf{Set}]$;
complete, cocomplete, locally small.

\medskip
\textbf{Simplicial identities.}\quad For $d_i = X(\delta^i)$, $s_i = X(\sigma^i)$:
\begin{align*}
  d_i d_j &= d_{j-1} d_i,            & i &< j; \\
  s_i s_j &= s_{j+1} s_i,            & i &\leq j; \\
  d_i s_j &= s_{j-1} d_i,            & i &< j; \\
  d_i s_j &= \mathrm{id},            & i &\in \{j, j+1\}; \\
  d_i s_j &= s_j d_{i-1},            & i &> j+1.
\end{align*}

\medskip
\textbf{Yoneda for $\mathbf{sSet}$.}\quad $\mathbf{sSet}(\Delta^n, X) \cong X_n$;
$\Delta^n_k = \Delta([k], [n])$.

\medskip
\textbf{Boundary and horn.}\quad
\begin{equation*}
  \partial\Delta^n \;=\; \bigcup_{i = 0}^{n} \delta^i(\Delta^{n-1}), \quad
  \Lambda^n_k \;=\; \bigcup_{i \neq k} \delta^i(\Delta^{n-1}).
\end{equation*}
Inner: $0 < k < n$; outer: $k \in \{0, n\}$.

\medskip
\textbf{Skeleta.}\quad $\mathrm{sk}_n \dashv \iota_n^* \dashv \mathrm{cosk}_n$
with $\iota_n : \Delta_{\leq n} \hookrightarrow \Delta$. Canonical
filtration $\mathrm{sk}_n X \hookrightarrow X$, $X = \mathrm{colim}_n
\mathrm{sk}_n X$.

\medskip
\textbf{Key simplicial sets.}\quad
\begin{itemize}
  \item $\Delta^n$ — representable, $n$-skeletal; terminal of $\mathbf{sSet}$ is $\Delta^0$.
  \item $\partial\Delta^n$ — boundary, $(n-1)$-skeletal.
  \item $\Lambda^n_k$ — horn, $(n-1)$-skeletal.
  \item $\mathrm{N}(\mathcal{C})$ — nerve; $\mathrm{N}(\mathcal{C})_n =$
        $n$-fold composable chains.
  \item $\mathrm{Sing}(X)$ — singular; $\mathrm{Sing}(X)_n =
        \mathbf{Top}(\lvert\Delta^n\rvert, X)$; always a Kan complex.
\end{itemize}
\end{formalrestatement}
